---
title: "Producing plots with `ggplot2`"
author: "Anne Vinkel"
date: "Statistical Programming with R"
output:
  ioslides_presentation:
    css: css/style.css
    logo: css/logo_en.png
    transition: 0
    self_contained: TRUE   
---

## Packages used

```{r, message=FALSE}
library(ggplot2)
library(mice) # for the boys dataset
library(tidyverse)
library(gridExtra) # to combine plots
```


# Background

## GGplot2

The `ggplot2` package by Hadley Wickham

 - produces layered statistical figures.

 - uses a “grammar” to build figures component-by-component
 
 - less immediately approachable than base `R` plots but much easier to customize
 
 - a part of the tidyverse
 
## Documentation 

https://ggplot2.tidyverse.org/reference/ is the officil homepage with help files and examples

The book  *ggplot2: Elegant Graphics for Data Analysis* by Hadley Wickham is a thorough and concise explanation, and is now available online: https://ggplot2-book.org/index.html

# The Grammar of Graphics

## The Grammar of Graphics

A concept by Leland Wilkinson (2005):

 - A grammar of a language defines the rules of structuring words and phrases into meaningful expressions

 - A grammar of graphics defines the rules of structuring mathematic and aesthetic elements into a meaningful graph

## Components of a ggplot 

- **Data**: which will be displayed on the plot
- **Aesthetic mappings**: How is data mapped to color, size, shape, location...
- **Geometries**: points, lines, shapes...
- **Stats**: stastical transformations - binning, smoothing, quantiles...
- **Scales**: Scales of aesthetic mappings (e.g. Danish=blue, Georgian=red)
- **Coordinate systems**: the plane on which data are mapped on the graphic.
- **Facets**: how should data be split into panels

# Construction of a ggplot

## Construction of a ggplot

Plots are built up in layers, generally in this order:

- Specify the data 
- Map variables to aesthetics
- Specify a plot
- Overlay summaries
- Add metadata, annotations

Data must be a dataframe (or tibble). Multiple dataframes can be plotted in the same ggplot


## Example - the `boys` data: scatterplot

Specify the data 
```{r, eval=FALSE}
ggplot(boys)
```

Map variables to aesthetics
```{r, eval=FALSE}
ggplot(boys, aes(x = age, y = bmi))
```

Specify a plot
```{r, eval=FALSE}
ggplot(boys, aes(x = age, y = bmi)) +  geom_point()
```

## An example: scatterplot
```{r, echo=FALSE, fig.align='center', message=F, warning=F}
boys %>%
  filter(!is.na(age) & !is.na(bmi)) %>% 
  ggplot(aes(x = age, y = bmi)) +
  geom_point() 
```

## Layers
New layers can be added to the plot
```{r, eval=F}
boys %>%
  filter(!is.na(age) & !is.na(bmi)) %>% 
  ggplot(aes(x = age, y = bmi)) +
  geom_point() +
  geom_smooth()
```

## Layers
```{r, echo=F, message=F}
boys %>%
  filter(!is.na(age) & !is.na(bmi)) %>% 
  ggplot(aes(x = age, y = bmi)) +
  geom_point() +
  geom_smooth()
```


## Aesthetics

- x
- y
- size
- colour
- fill
- opacity (alpha)
- linetype
- ...

## Aesthetics
```{r, warning=FALSE, message=FALSE, fig.align='center', dev.args=list(bg="transparent")}
gg <- 
  boys %>% 
  filter(!is.na(reg)) %>% 
  
  ggplot(aes(x      = age, 
             y      = bmi, 
             size   = hc, 
             colour = reg)) +
  
  geom_point(alpha = 0.5) +
  
  labs(title  = "BMI trend for boys",
       x      = "Age", 
       y      = "BMI", 
       size   = "Head circumference",
       colour = "Region") +
  theme_minimal()
```
 
## Aesthetics
```{r, warning=FALSE, message=FALSE, fig.align='center', dev.args=list(bg="transparent")}
plot(gg)
```

## Aesthetics

Default aesthetics are shared by all layers of the plot. You can do layer-specific aesthetics:

```{r, eval=F}
boys %>%
  ggplot( aes(x = age, y = bmi)) +
  geom_point( aes( col = reg )) + 
  geom_smooth()
```


## Aesthetics

```{r, echo=F}
boys %>% filter(!is.na(age) & !is.na(bmi) & !is.na(reg)) %>%
  ggplot( aes(x = age, y = bmi)) +
  geom_point( aes( col = reg )) + 
  geom_smooth()
```

## Mapping vs. setting aesthetics

**Map** variables to aesthetics *inside* the `aes` function, and the aesthetic varies with the variable. 

**Set** aesthetics to some constant level *outside* the aes function

```{r, eval=F}
boys %>%
  ggplot( aes(x = age, y = bmi)) +
  geom_point( aes( col = reg )) + 
  geom_smooth( col = "green")
```

## Geometries

Geometries are defined in part by which aesthetics they allow and require. You can find information on this in the help file for each geometry

- geom_point
- geom_bar
- geom_line
- geom_smooth

- geom_histogram
- geom_boxplot
- geom_density

## Geoms: Bar
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', dev.args=list(bg="transparent")}
data.frame(x = letters[1:5], y = c(1, 3, 3, 2, 1)) %>% 
  ggplot(aes(x = x, y = y)) + 
  geom_bar(fill = "dark green", stat = "identity") +
  labs(title = "Value per letter",
       x     = "Letter", 
       y     = "Value") +
  theme_minimal()
```

## Geoms: Line
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', dev.args=list(bg="transparent")}
ggdat <- data.frame(x = 1:100, y = rnorm(100))
ggdat %>% 
  ggplot(aes(x = x, y = y)) + 
  geom_line(colour = "dark green", lwd = 1) +
  ylim(-2, 3.5) +
  labs(title = "Some line thing",
       x     = "Time since start", 
       y     = "Some value") +
  theme_minimal()
```

## Geoms: Smooth
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', dev.args=list(bg="transparent")}
ggdat %>% 
  ggplot(aes(x = x, y = y)) + 
  geom_smooth(colour = "dark green", lwd = 1, se = FALSE) +
  ylim(-2, 3.5) +
  labs(title = "Some line thing",
       x     = "Time since start", 
       y     = "Some value") +
  theme_minimal()
```

## Geoms: Boxplot
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', dev.args=list(bg="transparent")}
boys %>% 
  filter(!is.na(reg)) %>% 
  
  ggplot(aes(x = reg, y = bmi, fill = reg)) +
  
  geom_boxplot() +
  
  labs(title = "BMI across regions",
       x     = "Region", 
       y     = "BMI") +
  theme_minimal() + 
  theme(legend.position = "none")
```

## Stats

Stat functions transform data, for example:

- frequencies of values of a variable (histogram, bar graphs)
- a mean
- a confidence limit

Each stat function has a default geom, so no geom is required for shapes to be rendered. The geom may be respecfied in the stat for different shapes.

## Stats

- `stat_bin` bins the data and by default produces a histogram
- `stat_summary` summarises y at each x (with a mean + se if nothing else is specified) and by default produces a pointrange-plot
- `stat_qq` produces quantile-quantile plots
- ...

## Stats

```{r, eval = F}
ageplot <- boys %>% 
  ggplot( aes( x = age)) +
  stat_bin()

ageplot2 <- boys %>% 
  ggplot( aes( x = age)) +
  stat_bin( geom = "line")

grid.arrange( ageplot, ageplot, ncol=2)
```


## Scales

Scales define which aesthetic values are mapped to data values - eg. if region is mapped to colour, which colour values are used for each region?

`scale_`-functions allow us to control the scales

```{r, eval=F}
boys %>%
  ggplot( aes(x = age, y = bmi, shape = reg )) +
  geom_point( ) +
  scale_shape_manual( values = c(5,3,8,13,20))
```

## Scales

```{r, echo=F, warning=F}
boys %>%
  ggplot( aes(x = age, y = bmi, shape = reg )) +
  geom_point( ) +
  scale_shape_manual( values = c(5,3,8,13,20))
```

## Colour scale functions

Colour is used a lot, and can be controlled with `scale_col_*`- functions and `scale_fill_*`-functions. For example:

- `scale_fill_hue()`: evenly spaced hues
- `scale_fill_brewer()`: sequential, diverging or qualitative colour schemes, originally intended to display factor levels on a map.
- `scale_fill_viridis_*`: colour maps that are colourblind-friendly and perceptually uniform in both colour and black-and-white


## Colour scale functions

```{r, eval=F}
dens <- boys %>% 
  mutate( agegrp = case_when( age < 10 ~ "<10",
                              age < 15 ~ "10-14",
                              age >= 15 ~ "15+")) %>%
  ggplot(aes(x=hgt, fill = agegrp))+
  geom_density( alpha=1/3)

grid.arrange( dens, dens + scale_fill_hue(), ncol=2)
```

## Colour scales
```{r, echo=F, warning=F, message=F}
dens <- boys %>% 
  mutate( agegrp = case_when( age < 10 ~ "<10",
                              age < 15 ~ "10-14",
                              age >= 15 ~ "15+")) %>%
  ggplot(aes(x=hgt, fill = agegrp))+
  geom_density( alpha=1/3)

grid.arrange( dens, dens + scale_fill_viridis_d(), ncol=2)
```

## Guides

Guides (axes and legends) visualise a scale 
- The x-axis visualises the mapping of data values to points on the x-axis
- The legend tells us which values map to which colours

Guides are displayed by default and can be manipulated with the `guides`-functions. (Eg. `guides(colour="none")` to remove colour-legend)

## Coordinate systems

Nearly always default cartesian, but can be (eg) polar, flipped cartesian, map projections...

## Faceting

Split plots into panels using `facet_wrap` and `facet_grid`

```{r, eval = F}
boys %>%
  ggplot( aes(x = age, y = bmi)) +
  geom_point( ) + 
  geom_smooth() + 
  facet_wrap( ~ reg)
```

## Faceting

```{r, echo = F, warning=FALSE, message=FALSE}
boys %>%
  filter( !is.na(age) & !is.na(bmi)) %>%
  ggplot( aes(x = age, y = bmi)) +
  geom_point( ) + 
  geom_smooth() + 
  facet_wrap( ~ reg)
```

## Themes

Themes control the elements of the graph not related to the data

- Background colour
- Font and font size
- Gridlines
- ...

You can apply an existing theme to your plot, or create your own

The `theme` functions allows us to adjust the theme

## Themes

```{r, eval = F}
boys %>%
  ggplot( aes(x = age, y = bmi)) +
  geom_point( ) +
  theme( panel.background = element_blank()
       , axis.title.x=element_text( size=14, color="red" ))
```

## Themes

```{r, echo = F, warning=F}
boys %>%
  ggplot( aes(x = age, y = bmi)) +
  geom_point( ) +
  theme( panel.background = element_blank()
       , axis.title.x=element_text( size=14, color="red" ))
```


## Saving plots

The `ggsave` functions makes saving plots easy

```{r, eval=F}
# Save as pdf
ggsave( "/output/plot.pdf", myplot)

# Save as png and resize
ggsave( "/output/plot.png", myplot, width=7, height=5, units="in")
```

# Practical 