---
title: "Practical tidy data"
author: "JPJ"
date: "15/10/2021"
output: 
  html_document:
    code_folding: hide
---

# Excercise

```{r message=FALSE, warning=FALSE, echo=F}

knitr::opts_chunk$set(message = F, results = F)

```

Inspiration from:  

* [R for Data Science: 12 Tidy Data](https://r4ds.had.co.nz/tidy-data.html#spreading-and-gathering)  
* [Tidy Data manifest](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html)  
* [r-exercises.com](https://www.r-exercises.com/2017/10/26/how-to-tidy-up-your-data-set-with-tidyr/?__cf_chl_managed_tk__=pmd_dWFMOu7CQXalYXDA4JRMEHfPY6or_CrfK2i91oS_fIo-1634307577-0-gqNtZGzNAyWjcnBszQjR)  
 


```{r class.source = 'fold-show', message=FALSE, warning=FALSE}

library(tidyverse)
```


1. **Use the tool box from the tidy-data session to make the data.tables below tidy**

1.a: **below is a table showing peoples age and height. Make the table tidy**
```{r class.source = 'fold-show', message=FALSE}


people <- tribble(
  ~name,             ~names,  ~values,
  "Phillip Woods",   "age",       45,
  "Phillip Woods",   "height",   186,
  "Jessica Cordero", "age",       37,
  "Jessica Cordero", "height",   156
)
```
**Hint** [press code to see the hint]{style="float:right"}
```{r}
#' Use pivot_wider
```



**Solution** [press code to see the hint]{style="float:right"}
```{r}
people %>% 
  pivot_wider(names_from = names, values_from = values)
```

1.b: **below is a table showing pregnancies on gender. Make the table tidy**
```{r class.source = 'fold-show', message=FALSE}
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes", NA, 10,
  "no", 20, 12
)
```


**Hint**
```{r}
#' Use pivot_longer
```

**Solution**
```{r}
preg %>% 
  pivot_longer(cols = c(male, female), names_to = "sex", values_to = "count")


# alternativ solution
preg %>% 
  pivot_longer(c(male, female), names_to = "sex", values_to = "count", values_drop_na = TRUE)
```



1.c: **below is a table showing income groups on religious belief. Make the table tidy**
```{r class.source = 'fold-show', results=T}
relig_income
```

**Solution**
```{r}
relig_income %>% 
  pivot_longer(cols = -religion,  names_to = "income", values_to = "frequency")
```


2. **Make sure that the data table billboard is tidy and compressed**
```{r class.source = 'fold-show', results=T}
billboard
```


**Hint**
```{r}

# Use pivot_longer for all weeks, compress the new variable from names_to and correct date.entered to first date of the week, call the new column 'date'. You should always find back to date.entered by taking min(date)
```

**Solution**
```{r}

billboard %>% 
  pivot_longer(
    wk1:wk76, 
    names_to = "week", 
    values_to = "rank", 
    values_drop_na = TRUE
  ) %>% 
  mutate(
    week = as.integer(str_remove(week, "wk")),
    date = as.Date(date.entered) + 7 * (week - 1),
    date.entered = NULL
  )

# Alternative solution

billboard %>% 
  pivot_longer(-c("artist", "track", "date.entered"),
    names_to = "week", 
    values_to = "rank", 
    values_drop_na = TRUE
  ) %>% 
  mutate(
    week = as.integer(str_remove(week, "wk")),
    date = as.Date(date.entered) + 7 * (week - 1),
    date.entered = NULL
  )
```


3. **Tidy the data below and calculate the sum and mean of all the points scored that day**
```{r class.source = 'fold-show'}
nba <- data.frame(
  player = c("James", "Durant", "Curry", "Harden", "Paul", "Wade"),
  team = c("CLEOH", "GSWOAK", "GSWOAK", "HOUTX", "HOUTX", "CLEOH"),
  day1points = c("25","23","30","41","26","20"),
  day2points = c("24","25","33","45","26","23")
)
```

**Hint1**
```{r}

#use pivot_longer on dayXpoints

```

**Hint2**
```{r}

# The first 3 character of the word is the team. The rest is the stat

```

**Solution (tidy)**
```{r}

nba_tidy <- nba %>% 
  pivot_longer(cols = c(day1points, day2points), names_to = "day", values_to = "points") %>%
  mutate(day = str_remove_all(day, "day|points") %>% as.numeric(),
         points = as.numeric(points)) %>% 
  separate(col = team, into = c("team", "state"), sep = 3) 

```

**Solution (with calculations)**
```{r}

nba_tidy %>% 
  group_by(day) %>% 
  summarise(total_points = sum(points),
            mean_points = mean(points),
            .groups = "drop")

```

