---
title: "Imputation"
author: "Jolien Cremers"
date: "Statistical Programming with R"
output:
  ioslides_presentation:
    css: css/style.css
    logo: css/logo_en.png
    transition: 0
    self_contained: TRUE   
---
# Working with distributions in `R`
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```

## Sampling from distributions:

In R we can use:

- `rnorm()` to obtain samples from a normal distribution
- `dnorm()` to sample from a normal density function
- `pnorm()` to sample from a normal distribution function
- `qnorm()` to sample from a normal quantile function

This works for other distributions as well, e.g. `rexp()`, `dexp()`, `pexp()` and `qexp()`.

## Setting a seed

```{r}
set.seed (101)
rnorm(5, 2, 5)
rnorm(5, 2, 5)

set.seed(101)
rnorm(5, 2, 5)
```

## Histograms

```{r echo=FALSE, fig.height=4}
set.seed(101)
tibble( x = rnorm(1000, 3, 0.5)) %>% 
  ggplot(aes( x = x))+
  geom_histogram( aes( y = ..density..)) + 
  geom_density( col = "red", size = 1) +
  ggtitle("Histogram of rnorm(1000,3,0.5")

```

## Code for previous plot

```{r, eval = FALSE, fig.height=4}
set.seed(101)
tibble( x = rnorm(1000, 3, 0.5)) %>% 
  ggplot(aes( x = x))+
  geom_histogram( aes( y = ..density..)) + 
  geom_density( col = "red", size = 1) +
  ggtitle("Histogram of rnorm(1000,3,0.5")
```

## Plotting distributions

Using `ggplot` with the `geom_function` geometry

```{r fig.height=3}
library(ggplot2)
ggplot() + xlim(c(-3,3)) + geom_function( fun = dnorm)
```

## Plotting distributions

Using `ggplot` with the `geom_function` geometry

```{r fig.height=3}
ggplot() + xlim(c(-3,3)) + geom_function( fun = pnorm)
```

# Imputation

## This Lecture:

- Missing data patterns
- Imputation methods
  * `simputation` package

## I use the following packages in this lecture

```{r message = FALSE}

library(naniar) # additional package for practical
library(simputation) # common interface for several imputation methods
library(tidyverse) 

```

# Missing data patterns

## `nhanes` data {.smaller}

`age`: Age group (1=20-39, 2=40-59, 3=60+)

`bmi`: Body mass index (kg/m**2)

`hyp`: High blood pressure, hypertension (1=no,2=yes)

`chl`: Total serum cholesterol (mg/dL)

```{r}
nhanes <- mice::nhanes

head(nhanes, n = 10)
```


## Exploring missingness patterns

First, a basic dplyr approach:
```{r}
nhanes %>% is.na() %>% colSums()
```
## Exploring missingness patterns

Two packages worth mentioning: `VIM` and `naniar` - we will work mainly with `naniar`

```{r}
miss_var_summary(nhanes)
```

## Visualise missing data pattern

```{r}
naniar::gg_miss_var(nhanes, show_pct = TRUE)
```

## Visualise missing data pattern

```{r}
naniar::gg_miss_upset(nhanes)
```

## Visualise missing data pattern by factor variable

```{r}
nhanes %>% mutate( age = factor(age)) %>%
  gg_miss_fct(fct = age)
```


## Visualization
```{r, message = FALSE, warning = FALSE}
naniar::gg_miss_fct(x = nhanes, fct = age)
```


## Visualization
```{r}
vis_miss(nhanes, cluster= T)
```

# Imputation methods

## Specifying models in `simputation`

Common structure for specifying an imputation model:

```{r, eval = FALSE}
impute_<model>(data, formula, [model-specific options])
```

Structure of the model in `formula`:

```{r, eval = FALSE}
IMPUTED ~ MODEL_SPECIFICATION [ | GROUPING ]
```


## Regression imputation {.smaller}

```{r}
imp_1 <- impute_lm(nhanes, bmi ~ age + hyp + chl)
head(imp_1)
head(nhanes)
```

## Median imputation

```{r}
imp_2 <- impute_median(imp_1, bmi ~ age)
head(imp_2)
```

##

```{r}

imp_2a <- nhanes %>%
          impute_lm(bmi ~ age + hyp + chl) %>%
          impute_median(bmi ~ age)
head(imp_2a)
```

## Other methods

- `impute_cart()`: decision tree imputation
- `impute_hotdeck()`:  random and sequential hot deck, k-nearest neighbours imputation and predictive mean matching.
- `impute_multivariate()`: imputation based on EM-estimation of multivariate normal parameters, imputation based on iterative Random Forest estimates and stochastic imptuation based on bootstrapped EM-estimatin of multivariate normal parameters.
- `impute_proxy()`: Impute missing values by a constant, by copying another variable, computing transformations from other variables.

# Practical 
