---
title: "Practical D"
author: "Jolien Cremers + Anne Vinkel"
date: "Statistical Programming with R"
output: html_document
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
printAnswers <- TRUE
```

----

The purpose of this practical is to use some of R's packages to solve a missing data problem by imputing missing values in a 'real' dataset. This can either be a dataset you've brought to this course yourself or the `selfreport` data from the `mice` package. The 'selfreport' data is described in Section 7.3 of the first edition of the book Flexible Imputation of Missing Data (van Buuren, S. (2012). *Flexible Imputation of Missing Data.* CRC/Chapman & Hall, FL: Boca Raton).

---

Suggested packages:

```{r message = FALSE}
library(naniar) # additional package for practical
library(simputation) # common interface for several imputation methods
library(tidyverse)
```
**Note that:** A potential issue is that some of the functions in `simputation` and `naniar` have the same names - e.g. `impute_median`. This means that the order in which the packages are loaded matters - the last loaded package masks the functions from the former package. To specify which package you mean, call `simputation::impute_median` and `naniar::impute_median`

---

#### Selfreport data
```{r}
selfreport <- mice::selfreport
```

---

The selfreport data contains data from two studies, one from the `mgg` study, the *survey dataset*, where height and weight are selfreported measures (`hr` and `wr`) and one from the `krul` study, the *calibration dataset*, where the height and weight measures are both selfreported and measured (variables `hm` and `wm`). The goal for the self-report data is to obtain correct estimates of the prevalence of obesity (Body Mass Index > 32) in the *survey dataset*. Note that the Body Mass Index (bmi) is computed as follows:

$$\text{bmi} = \frac{\text{weight in kg}}{(\text{height in m})^2} $$

---

1. **Check missing data patterns.**

First, check if there is a description of the `selfreport` data in the `mice` package documentation:

```{r}
?selfreport
```

The selfreport data contains data from two studies, one from the `mgg` study where height and weight are selfreported measures (`hr` and `wr`) and one from the `krul` study where the measures are both selfreported and measured (variables `hm` and `wm`). 

Then, investigate the missing data patterns

```{r}
head(selfreport, n = 10)
```

```{r}
summary(selfreport)
```

The measures weight and height variables are indeed missing from the `mgg` study

```{r}
gg_miss_fct(selfreport, src)
gg_miss_var(selfreport, show_pct = TRUE)
```

Apart from measured height and weight also the measured bmi, the indicators for pregnancy, educational level and ethnicity contain missing values.

```{r}
gg_miss_upset(selfreport, nsets = 15)
```

---

2. **Check the relation between measured and self reported bmi in the *calibration dataset*.**

There are more individuals that have a lower selfreported BMI than measured BMI.

```{r}

calibration <- selfreport %>% 
  filter(as.numeric(src) == 1) %>% 
  mutate(difference = case_when(br < bm ~ "lower",
                                bm < br ~ "higher",
                                bm == br ~ "same"))

calibration %>% count(difference)

ggplot(calibration, aes(br, bm, color = difference)) +
  geom_point() +
  theme_minimal() +
  geom_abline(intercept = 0, slope = 1) + 
  labs(x = "Selfreported BMI", y = "Measured BMI", fill = "")

```

---

3. **Impute missing data using the simputation package - try, for instance impute_rf**

First define a deterministic function to compute bmi:

```{r}
bmi <- function(h, w) {
  return(w/(h/100)^2)
}
```

One approach could be `impute_rf` which implements a random forest to do the imputation

```{r, message=FALSE}
# Impute prg
selfreport %>% count(sex, prg)

# We assume that missingness of prg implies "not pregnant"
selfreport <- selfreport %>%
  mutate( prg = replace_na( as.character(prg), "not pregnant")
        , prg = as.factor(prg)) %>% 
  # The factor etn has missing levels that will cause trouble
  mutate( etn = droplevels(etn))

# First impute measured height and weight
library(randomForest)
selfreport_imp_rf <- impute_rf( selfreport,
                                formula = hm + wm ~ hr + wr + web  + src
                                              + age + sex + prg)

# Compute imputed bmi from imputed height and weight
selfreport_imp_rf <- selfreport_imp_rf %>%
  mutate( bm = bmi(hm,wm))
head(selfreport)
head(selfreport_imp_rf)

# Impute edu and eth
selfreport_imp_rf <- impute_rf( selfreport_imp_rf,
                                formula = edu + etn ~ hr + wr + hm + wm + bm + br
                                          + web + src + age + sex + prg)

tail(selfreport)
tail(selfreport_imp_rf)

any_miss(selfreport_imp_rf)
```


---

4. **What is the estimated prevalence (selfreported and imputed data) of obesity (defined as a measured bmi of 32 or above) for different `age` and `sex` in the *survey dataset*?**

Estimated prevalence in the selfreported data

```{r}

selfreport_imp_rf %>% 
  filter( src == "krul") %>%
  mutate(bm = bmi(hm, wm)) %>%
  mutate(obese = case_when(br > 32 ~ 1,
                           TRUE ~ 0)) %>% 
  mutate(age_cat = cut(age, c(17, 29, 39, 49, 59, 75))) %>% 
  group_by(src, sex, age_cat) %>% 
  summarise(N = n(), prevalence = sum(obese)/N*100)

```

Estimated prevalence in the imputed data

```{r}
selfreport_imp_rf %>% 
  filter(src=="mgg") %>% 
  mutate(bm = bmi(hm, wm)) %>%
  mutate(obese = case_when(br > 32 ~ 1,
                           TRUE ~ 0)) %>% 
  mutate(age_cat = cut(age, c(17, 29, 39, 49, 59, 75))) %>% 
  group_by(src, sex, age_cat) %>% 
  summarise(N = n(), prevalence = sum(obese)/N*100)

```

---

End of practical. 
