---
title: "Practical D"
author: "Jolien Cremers + Anne Vinkel"
date: "Statistical Programming with R"
output: html_document
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
printAnswers <- TRUE
```

----

The purpose of this practical is to use some of R's packages to solve a missing data problem by imputing missing values in a 'real' dataset. This can either be a dataset you've brought to this course yourself or the `selfreport` data from the `mice` package. The 'selfreport' data is described in Section 7.3 of the first edition of the book Flexible Imputation of Missing Data (van Buuren, S. (2012). *Flexible Imputation of Missing Data.* CRC/Chapman & Hall, FL: Boca Raton).

---

Suggested packages:

```{r message = FALSE}
library(naniar) # additional package for practical
library(simputation) # common interface for several imputation methods
library(tidyverse)
```
**Note that:** A potential issue is that some of the functions in `simputation` and `naniar` have the same names - e.g. `impute_median`. This means that the order in which the packages are loaded matters - the last loaded package masks the functions from the former package. To specify which package you mean, call `simputation::impute_median` and `naniar::impute_median`

---

#### Selfreport data
```{r}
selfreport <- mice::selfreport
```

---

The selfreport data contains data from two studies, one from the `mgg` study, the *survey dataset*, where height and weight are selfreported measures (`hr` and `wr`) and one from the `krul` study, the *calibration dataset*, where the height and weight measures are both selfreported and measured (variables `hm` and `wm`). The goal for the self-report data is to obtain correct estimates of the prevalence of obesity (Body Mass Index > 32) in the *survey dataset*. Note that the Body Mass Index (bmi) is computed as follows:

$$\text{bmi} = \frac{\text{weight in kg}}{(\text{height in m})^2} $$

---

1. **Check missing data patterns.**

**Hint:** First, check if there is a description of the `selfreport` data in the `mice` package documentation:

---

2. **Check the relation between measured and self reported bmi in the *calibration dataset* (src=="krul")**


---

3. **Impute missing data using the simputation package - try, for instance impute_rf**

---

4. **What is the estimated prevalence (selfreported and imputed data) of obesity (defined as a bmi of 32 or above) for different `age` and `sex` in the *survey dataset*?**

---

End of practical. 
